Appendix

### 2025-02-18 manuscript_rmarkdown/manuscript_appendix.Rmd
Add table with p-values for ecn test.

Table \ref{tab:normalisation-before-transformation-appendix} shows residual errors after transformation to normality of the five features from real-world datasets presented previously in Appendix D and the main manuscript.
Table \ref{tab:normalisation-before-transformation-lambda-appendix} shows the corresponding values for transformation parameter $\lambda$.
Normalisation prior to transformation prevents unexpectedly high $\lambda$ values in the presence of large location shifts, as can be observed for the latitude feature.
Normalisation prior to transformation, however, does not consistently improve residual errors after transformation. 
Interestingly, use of normalisation prior to transformation with robust power transformation using the method by Raymaekers and Rousseeuw [@Raymaekers2024-zf] led to higher residual errors for arterial wall thickness, fuel efficiency and latitude features.
This is similar to observations by @Raymaekers2024-zf, who observed that their method can emphasize values of outliers after transformation.
Location- and scale-invariant transformations have similar or lower residual errors compared to errors resulting from normalisation prior to transformation.

@@

data <- .assess_standardisation_before_transformation()
data <- data[, mget(c("residuals", "lambda", "transformation", "dataset"))]
data[, "residuals" := round(residuals, digits = 1L)]
data[, "lambda" := round(lambda, digits = 1L)]
data.table::dcast(data, transformation ~ dataset, value.var = "residuals")
data.table::dcast(data, transformation ~ dataset, value.var = "lambda")
```

```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from real-world datasets after Yeo-Johnson transformation to normality. AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-lambda-appendix}
\begin{tabular}{l | c c c c c}

@@

>>>>

Table \ref{tab:normalisation-before-transformation-appendix} shows residual errors after transformation to normality of the five features from real-world datasets presented previously in Appendix D and the main manuscript.
Table \ref{tab:normalisation-before-transformation-lambda-appendix} shows the corresponding values for transformation parameter $\lambda$, 
\ref{tab:normalisation-before-transformation-p-value-appendix} shows p-values of the empirical central normality test.
Normalisation prior to transformation prevents unexpectedly high $\lambda$ values in the presence of large location shifts, as can be observed for the latitude feature.
Normalisation prior to transformation, however, does not consistently improve residual errors after transformation. 
Interestingly, use of normalisation prior to transformation with robust power transformation using the method by Raymaekers and Rousseeuw [@Raymaekers2024-zf] led to higher residual errors for arterial wall thickness, fuel efficiency and latitude features.
This is similar to observations by @Raymaekers2024-zf, who observed that their method can emphasize values of outliers after transformation.
Location- and scale-invariant transformations have similar or lower residual errors compared to errors resulting from normalisation prior to transformation.
Not surprisingly, normalisation before transformation does not noticeably improve central normality, as assessed using the empirical central normality test.

@@

data <- .assess_standardisation_before_transformation()
data <- data[, mget(c("residuals", "lambda", "p_value", "transformation", "dataset"))]
data[, "residuals" := round(residuals, digits = 1L)]
data[, "lambda" := round(lambda, digits = 1L)]
data[, "p_value" := round(p_value, digits = 3L)]
data.table::dcast(data, transformation ~ dataset, value.var = "residuals")
data.table::dcast(data, transformation ~ dataset, value.var = "lambda")
data.table::dcast(data, transformation ~ dataset, value.var = "p_value")
```

```{=latex}
\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-lambda-appendix}
\begin{tabular}{l | c c c c c}

@@

```{=latex}
\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from real-world datasets after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-p-value-appendix}
\begin{tabular}{l | c c c c c}

\midrule
none & 0.685 & 0.558 & 0.764 & 0.354 & 0.030 \\
conventional & 0.886 & 0.010 & 0.012 & 0.317 & 0.381 \\
conventional (z-score norm.) & 0.842 & 0.036 & 0.011 & 0.075 & 0.378 \\
conventional (robust scaling) & 0.855 & 0.002 & 0.007 & 0.077 & 0.416 \\
Raymaekers-Rousseeuw & 0.886 & 0.610 & 0.553 & 0.296 & 0.381 \\
Raymaekers-Rousseeuw (z-score norm.) & 0.685 & 0.632 & 0.812 & 0.075 & 0.378 \\
Raymaekers-Rousseeuw (robust scaling) & 0.826 & 0.609 & 0.813 & 0.077 & 0.434 \\
invariant & 0.907 & 0.134 & 0.026 & 0.049 & 0.549 \\
robust invariant & 0.781 & 0.102 & 0.583 & 0.085 & 0.717 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
```

### 2025-02-18 manuscript_rmarkdown/manuscript_appendix.tex
Add table with p-values for ecn test.

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2024-10-17}

@@

An alternative to location- and scale-invariant transformations is
normalising feature distributions prior to conventional transformations.
Table \ref{tab:normalisation-before-transformation-appendix} shows
residual errors, after transformation to normality, of the five features
from real-world datasets presented previously in Appendix D and the main
manuscript. In these examples location- and scale-invariant
transformations have similar or lower residual errors compared to errors
resulting from normalisation prior to transformation.

\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. conv.: conventional; norm.: normalisation; rob.: robust}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & none & conventional & conv. (z-score norm.) & conv. (rob. scaling) & invariant \\

\midrule
age                     &  16.5 &  11.5 &  11.5 &  11.3 &   8.8 \\
penguin body mass       &  48.0 &  32.2 &  33.3 &  32.2 &  26.8 \\
latitude                & 328.1 & 319.0 & 326.2 & 324.5 & 326.4 \\
fuel efficiency         &  54.5 &  55.3 &  49.0 &  53.3 &  44.0 \\
arterial wall thickness & 110.1 &  30.0 &  19.3 &  31.8 &  12.2 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

>>>>

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-02-18}

@@

An alternative to location- and scale-invariant transformations is
normalising feature distributions prior to conventional transformations.
We assessed two normalisation methods were used: 1. z-standardisation,
i.e.~\(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
and \(\sigma\) the mean and standard deviation of feature
\(\mathbf{X}\); and 2. robust scaling,
i.e.~\(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / \text{IQR}\left(\mathbf{X}\right)\),
with \(\text{IQR}\) representing the interquartile range of feature
\(\mathbf{X}\).

Table \ref{tab:normalisation-before-transformation-appendix} shows
residual errors after transformation to normality of the five features
from real-world datasets presented previously in Appendix D and the main
manuscript. Table
\ref{tab:normalisation-before-transformation-lambda-appendix} shows the
corresponding values for transformation parameter \(\lambda\),
\ref{tab:normalisation-before-transformation-p-value-appendix} shows
p-values of the empirical central normality test. Normalisation prior to
transformation prevents unexpectedly high \(\lambda\) values in the
presence of large location shifts, as can be observed for the latitude
feature. Normalisation prior to transformation, however, does not
consistently improve residual errors after transformation.
Interestingly, use of normalisation prior to transformation with robust
power transformation using the method by Raymaekers and Rousseeuw
(Raymaekers and Rousseeuw 2024) led to higher residual errors for
arterial wall thickness, fuel efficiency and latitude features. This is
similar to observations by Raymaekers and Rousseeuw (2024), who observed
that their method can emphasize values of outliers after transformation.
Location- and scale-invariant transformations have similar or lower
residual errors compared to errors resulting from normalisation prior to
transformation. Not surprisingly, normalisation before transformation
does not noticeably improve central normality, as assessed using the
empirical central normality test.

\begin{table}
\begin{center}
\caption{Residual errors for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & age & AWT & FE & latitude & PBM \\

\midrule
none & 16.5 & 110.1 & 54.5 & 328.4 & 48.0 \\
conventional & 11.5 & 30.0 & 55.3 & 319.0 & 32.2 \\
conventional (z-score norm.) & 11.5 & 19.3 & 49.0 & 326.2 & 33.3 \\
conventional (robust scaling) & 11.3 & 31.8 & 53.3 & 324.5 & 32.2 \\
Raymaekers-Rousseeuw & 11.5 & 136.7 & 47.7 & 315.1 & 32.2 \\
Raymaekers-Rousseeuw (z-score norm.) & 13.2 & 214.5 & 57.4 & 326.2 & 33.3 \\
Raymaekers-Rousseeuw (robust scaling) & 11.7 & 281.3 & 57.0 & 324.5 & 32.0 \\
invariant & 8.8 & 12.2 & 44.0 & 326.4 & 26.8 \\
robust invariant & 10.6 & 27.1 & 49.5 & 310.6 & 23.1 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Transformation parameter $\lambda$ for features from real-world datasets after Yeo-Johnson transformation to normality. 
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-lambda-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & age & AWT & FE & latitude & PBM \\

\midrule
conventional & 2.0 & -0.7 & -0.1 & 62.1 & -0.5 \\
conventional (z-score norm.) & 1.2 & -1.7 & -0.1 & 1.3 & 0.6 \\
conventional (robust scaling) & 1.3 & 0.0 & 0.2 & 1.4 & 0.4 \\
Raymaekers-Rousseeuw & 2.0 & 1.1 & 0.8 & 95.4 & -0.5 \\
Raymaekers-Rousseeuw (z-score norm.) & 1.3 & 1.4 & 1.0 & 1.3 & 0.6 \\
Raymaekers-Rousseeuw (robust scaling) & 1.4 & 1.3 & 1.0 & 1.4 & 0.4 \\
invariant & 1.3 & 0.2 & -1.3 & 1.5 & 0.5 \\
robust invariant & 1.4 & -0.6 & 0.9 & 1.4 & 0.4 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{
p-values of empirical central normality tests for features from real-world datasets after Yeo-Johnson transformation to normality.
Several transformation methods include normalisation before tranformation, indicated by z-score normalisation (norm.) or robust scaling. 
AWT: arterial wall thickness; FE: fuel efficiency; PBM: penguin body mass}
\label{tab:normalisation-before-transformation-p-value-appendix}
\begin{tabular}{l | c c c c c}

\toprule
feature & age & AWT & FE & latitude & PBM \\

\midrule
none & 0.685 & 0.558 & 0.764 & 0.354 & 0.030 \\
conventional & 0.886 & 0.010 & 0.012 & 0.317 & 0.381 \\
conventional (z-score norm.) & 0.842 & 0.036 & 0.011 & 0.075 & 0.378 \\
conventional (robust scaling) & 0.855 & 0.002 & 0.007 & 0.077 & 0.416 \\
Raymaekers-Rousseeuw & 0.886 & 0.610 & 0.553 & 0.296 & 0.381 \\
Raymaekers-Rousseeuw (z-score norm.) & 0.685 & 0.632 & 0.812 & 0.075 & 0.378 \\
Raymaekers-Rousseeuw (robust scaling) & 0.826 & 0.609 & 0.813 & 0.077 & 0.434 \\
invariant & 0.907 & 0.134 & 0.026 & 0.049 & 0.549 \\
robust invariant & 0.781 & 0.102 & 0.583 & 0.085 & 0.717 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}
